### 2015-09-02 aphids/aphids.tex
updated aphids chapter

To enbable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
the data before correlating.  In this chapter we describe a software solution to change the 
samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
focus on the rate conversion problem and investigate several possible algorithms.  The project
has applications for future VLBI observations at the SMA as well as other sites with similar 
challenges.

@@

Very Long Baseline Interferometry (VLBI) uses similar techniques as the connected element arrays
featured in the previous chapters.  By using detectors located thousands of miles away, perhaps on different 
continents or even in space, this technique can achieve milliarcsecond resolution at radio wavelengths.  Each 
station must have its own clock and, after the observation campaign, the recorded signals is correlated off-site. 
Examples of VLBI applications include geodesy, astrometry, and imaging. Technical developments in the 
field continue to unveil singular scientific opportunities.

One exciting example is the development of mm and sub-mm wave VLBI by collaborations such as the Event Horizon
Telescope (EHT).  The EHT observes nearby super massive black holes on scales of their Schwarzschild radius which
provides a unique lab to study 
general relativity and accretion physics.  One of the challenges that this experiment faces is the inhomogeneity 
of its array since it is a conglomeration of many independently operating observatories.
While many of these telescopes use the same back end hardware such as the state-of-the-art 
ROACH2 VLBI Digital Back-end (R2DBE), a few sites 
cannot use these systems and produce very 
different data products. In particular, the sampling rate of the phased ALMA, SMA, and CARMA arrays are 
incompatible with the standard VLBI rate.  Before the EHT's software correlator
\citep[DIFX;][]{deller07} can find fringes using these data, they must be resampled to the R2DBE rate of 
4096\,MHz.

This chapter summarizes the Adaptive Phased-array and Heterogeneous 
Interpolator, and Downsampler for SWARM (APHIDS), a resampling software designed for 
phased SMA data products.  We concentrate our discussion to early development and design in the project with 
a particular emphasis on the resampling problem.  APHIDS also has very general applications for 
post-processing of VLBI data since it features fast read/write capabilities, a modular threaded pipeline 
and a GPU enabled fast data processing.  The code base is open-source and available online at 
\url{https://github.com/sma-wideband/sdbe}.

The purpose of APHIDS is to take data recorded at the SMA and process it for correlation using
DIFX.  The SMA is a key site for the EHT that provides important East-West baselines and a zero baseline with 
its neighbor JCMT.  The phased sum is calculated by the SMA correlator, the SMA Wideband Astronomical 
ROACH2 Machine (SWARM\footnote{\url{https://www.cfa.harvard.edu/twiki5/view/SMAwideband}}),
which is currently being deployed in stages as the array's bandwidth capabilities are 
upgraded.  For EHT observations in 2015, SWARM recorded at a sampling rate of 2496\,MHz in two separate phased
sums.

In \S\ref{sec:src_basics} we discuss the sampling rate conversion problem and several common solutions.
The current APHIDS resampling algorithm is described in \S\ref{sec:aphids} along with discussions of important
design considerations.  We include measurements of performance as well as a successful
demonstration of the software's operation.  We end by summarizing the goals for near and long-term project 
development.

Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  Operationally, 
sample rate conversions is the combination of upsampling and downsampling \citep{oppenheim10,lyons11}.  
Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
time and frequency domain for when $L$ equals 3.

@@

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}
   \label{fig:resample_basic}
   \node[dspnodeopen,dsp/label=left]  (c0) {$x[i]$};
   \node[dspsquare,right=of c0]                     (c1) {\upsamplertext{L}};
   \node[dspfilter,right=of c1]                     (c2) {$H(\z)$};
   \node[dspsquare,right=of c2]                    (c3) {\downsamplertext{M}};
   \node[dspnodeopen,right=of c3,dsp/label=right]  (c4) {$y[j]$};
   \foreach \i [evaluate = \i as \j using int(\i+1)] in {0,1,...,3}
       \draw[dspconn] (c\i) -- (c\j);
\end{tikzpicture}
\caption{Basic multirate resampling signal flow graph.}
\end{figure}

There have been recent development of polyphase filter representations for GPU architectures 
\citep[i.e.][]{vanderveldt12,adamek14,kim14a}.  These filter structures can operate at the low sample rate of 
$f_0/M$
by introducing delays to switch the expander and decimatator \citep{crochiere81}.  However, they require 
some filter design and careful book-keeping or buffering to keep track of sample indices \citep{wang01}.
The implementation in \citet{kim14a} assigned each thread to a filter coefficient, parallelizing the 
inner product operation.  However, \citet{kim14a} found that their GPU kernel was dominated by the 
indexing operations.  \citet{adamek14} (building on the work by \citet{vanderveldt12}) presented tantalizing 
implementations of 
polyphase filter banks that could run at data rates in excess of 6.5\,GB/s (their estimate for the single channel 
output of the SKA's Low Frequency Aperture Array).

%  https://github.com/wesarmour/astro-accelerate

A simple option for resampling is to use linear interpolation:
\begin{equation}
y[i] = x[j] + (x[j+1] - x[j]) (if_0/f_1 - j)
\end{equation}
where $j = \mathrm{floor}(if_0/f_1)$.  This is equivalent to applying a ``tent'' FIR filter with $2L$ taps on the 
upsampled $Lf_0$ time series before decimation \citep{oppenheim10}.  Similarly, a $2L$ tap box-car filter is 
equivalent to nearest-neighbor interpolation:
\begin{equation}
y[i] = x[\mathrm{round}(if_0/f_1)].
\end{equation}

@@

\begin{figure}[t!]
\epsscale{1.0}
\plotone{aphids/windows.eps}
\caption{Frequency response for FIR filters equivalent to linear and nearest interpolation when $L/M = 64/39$.
The orange region shows the first Nyquist zone of the target $L/Mf_0$ sample rate.  Spectral components at all 
other frequencies are aliased into this region.}
\label{fig:windows}
\end{figure}

\subsection{Resampling in the Fourier Domain} \label{sec:fourier_resampling}
In contrast to the previous methods, one can also implement a rational $L/M$ sample rate conversion entirely in 
the Fourier domain \citep{gold69,yeh82}.  After accumulating $kM$ samples at a clock rate of $f_0$ 
(where $k=1,2,3\cdots$), the DFT 
returns spectral components spaced at $f_0/kM$.  If $f_1 > f_0$, the resampled spectrum is generated by inserting
$p$ zeros to match the new $f_1$ while maintaining the correct frequency components: 
\begin{equation}
\frac{f_1}{kM+p} = \frac{f_0}{kM}.
\end{equation} 
Solving for $p = k(L - M)$.  If $f_0 < f_1$, the spectrum is instead trimmed by $p$ samples.  The time series
sampled at $f_1$ can then be constructed from the inverse DFT.  This method is equivalent to sinc 
interpolation using an ideal low-pass filter and is a perfect interpolator
if the original, continuous signal has a discrete spectral density distribution that is band-limited below the 
Nyquist limit.  In practice, small errors may be introduced from aliased spectral leakage.

Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
resampled signal.  The far this error significantly propagates away from the time series edges depends on the 
width of the sinc function which is set by the resampling factors $L$ and $M$: 
\begin{equation} \label{eq:sinc}
B\,\mathrm{sinc}(B t) \overset{\mathcal{F}}{\Longleftrightarrow} \begin{cases} 1 \quad |f| < B/2 \\ 0 \quad |f| > B/2 \end{cases}
\end{equation}
where $B = \mathrm{min}(f_0,L/Mf_0)$.  
The number of samples for which this makes a large difference is small and can be roughly treated as less than a 
1\% effect when $N > 100\,\mathrm{max}(L/M,1)$.  One option to mitigate this error to stitch together 
overlapping, resampled segments \citep{bi11} which requires more computation and memory.  A second strategy is to 
multiply the original signal by some tapered window function \citep{fraser89}.

For post-processing phased SWARM, DFT resampling is a good fit because one can access arbitrary large chunks of 
the time series without accumulators in hardware and has well-behaved errors
that can be easily modeled.  However, the speed of the FFT depends on the resampling factors ($L/M = 32/39$ for 
the March 2015 EHT observations) and zero-padding in the case that the $L\gg M$ may require large amounts of 
memory.

@@

@@

We are developing a software solution to the resampling problem called APHIDS (Adaptive 
Phased-array and 
Heteregeneous Interpolator and Downsampler for SWARM).  This open-source project is written in C/CUDA using
the High Availibility Shared Pipeline 
(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
\ref{fig:aphids_flow_chart} shows how data progresses 
through the pipeline and this section will describe each block covering the major hardware components, software 
design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.

\begin{figure}[t]
\linespread{1.}\selectfont{}
\begin{center}
\begin{tikzpicture} [node distance = 5cm, auto]
\node [cpu] (mark6_in) {Mark6 Reader};
\node[gpu,below of=mark6_in](reader){Reader};
\node[gpu,right of=reader](resampler){Resampler};
\node[cpu,right of=resampler](VDIF_pkt){VDIF packet \\ \& header};
\node[cpu,below of=VDIF_pkt](mark6_out){Mark6 Writer};

\path[line] (mark6_in) -- node[align=center] {BENG @ 2496 GHz \\ quantized to 2-bits}(reader);
\path[line] (reader) -- node[align=center] {spectra of \\ 16k complex64} (resampler);
\path[line] (resampler) --  node[align=center] {time series \\ @ 2048 GHz}(VDIF_pkt);
\path[line] (VDIF_pkt) -- node[align=center, left]{VDIF @ 4096 GHz \\ quantized to 2-bits}(mark6_out);
\end{tikzpicture}
\label{fig:aphids_flow_chart}
\caption{This signal flow chart shows an overview of APHIDS.  The red boxes represent operations on the GPU 
while blue boxes are for the CPU.  Once the B-engine frames have been unpacked, each of the two phased sum 
streams are processed independently and in a serial fashion.}
\end{center}
\end{figure}

@@

APHIDS features three main hardware components: an I/O data system, CPU, and GPU server.  For I/O, the 
EHT observations taken by SWARM in March, 2015 have been stored on four disk-packs that each hold eight, 
6\,TB hard drives (totaling $48 \times 4 = 192$\,TB of data).  The Mark6 VLBI data system has four modules that 
can hold one disk-pack and we use
two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
that controls the GPU server.

The GPU server as four GTX 980 GPUs, each with 4GB of local memory.  Data from the CPU (or host) is copied over 
a PCI-E bus into the local GPU (or device) memory.  Once the data is transferred, the host directs the device
to perform functions (or kernels) on the data which must be again moved from the local device memory onto 
on-chip memory for the individual streaming multiprocessors that make up the GPU.  After computation is done, 
the host transfers the data off of the device memory.

The observations were recorded using a scatter/gather file system which disperses the 
data composing individual scans over the four modules.  APHIDS uses fast, general purpose software called 
sgcomm (Scatter/Gather Communication) to read in 
the scattered data into CPU shared memory over a 10 Gigabit Ethernet network.  Data from phased SWARM was 
streamed through a special version of the R2DBE called the SDBE (Swarm Digital Back End) and packaged into
a customized B-engine format.  Each B-engine frame is composed of 1024, 1056B VDIF frames 
\footnote{\url{http://www.vlbi.org/vdif/}} with 128 snapshots of a 
16384 sample SWARM spectra.  However, the current SDBE bitcode packs
neither the snapshots contiguously in time nor the spectrum samples contiguously in frequency.  Furthermore, 
each B-frames overlaps the preceding one in memory.  Our resampling algorithm handles $19968$\,SWARM samples
simultaneously, so in order to reduce the amount of book-keeping the inout thread buffers 40 B-engine frames to 
produce $39 \times 128$ complete SWARM snapshots.  The inout thread does preliminary checks to ensure that the 
B-engine frames are complete before assigning each buffer to a GPU.  The data is written as a time series using 
following the VDIF specification.

As mentioned previously, the data is split into two phased sums that together cover the desired 4096\,MHz of 
observing bandwidth.  Since DIFX can handle the chunks independently we do not stitch the spectra together. 
However, the observation was designed to include guard bands which must be removed by APHIDS 
(see Figure \ref{fig:swarm_amp_spec}).

\begin{figure}[t!]
\epsscale{1.0}
\plotone{aphids/SWARM_amp_spectra.eps}
% /home/krosenfe/sdbe/software/prototyping/cuda/SWARM_amp_spectra.eps
\caption{This figure shows both channels of the detrended amplitude spectral density of phased SWARM along with 
the R2DBE (SMA single-dish in gray).}
\label{fig:swarm_amp_spec}
\end{figure}

@@

The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performace, we utilize batched
FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
and ensure that transforms operate on contiguous chunks of memory.  We also use single precision transforms to 
reduce the memory transfer bandwidth and ensure that 
the resampling block can be run on a single device.  The latter point is important since it allows the inout
thread to parallelize the resampling operation across the cluster and provide a roughly x4 speedup.

\begin{figure}[t!]
\linespread{1.}\selectfont{}
\centering
\begin{tikzpicture}[node distance = 2.2cm, auto]
\node[align=center](init){};
\node[resamp,below of=init](rfft){batched real FFT \\ dimension of $39 \times N$};
\node[resamp,below of=rfft](trim){Trim band and shift frequency};
\node[resamp,below of=trim](rifft){batched real IFFT \\ dimension of $32 \times N$};
\node[align=center,below of=rifft](final){};

\path[line](init) -- node[align=left]{ time series \\ @ 2496\,MHz} (rfft);
\path[line](rfft) -- (trim);
\path[line](trim) -- (rifft);
\path[line](rifft) -- node[align=left]{ time series \\ @ 2048\,MHz}(final);
\end{tikzpicture}
\label{fig:resampling_block}
\caption{The resampling block trims the phased SWARM guard bands and shifts the passband to DC as part of the 
resampling operation.}
\end{figure}

The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
32768 real samples.  Since SWARM omits the Nyquist sample which is far outside the SWARM cutoff frequency, 
we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
the Cooley-Tukey algorithm.  However, on our cluster the cuFFT library exhibits linear improvement in Gflops 
with transform size up to N of 16384 (see Figure \ref{fig:C2R_performance}).  This behavior is consistent with 
the CUDA 7.0 performance 
report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.

@@

The choice of FFT size was set to maximize performance.  In order for the target sampling rate to be achieved
$N_B$ must be a multiple of 39.  Furthermore, so that the 150\,MHz guard band can be removed, we must ensure that 
150 is a multiple of $2496/N_B$:
\begin{equation}
\frac{150\times39}{2496} = \frac{75}{32} = \frac{i}{j}, \quad i,j = 1, 2,3, \ldots
\end{equation}
To ensure this, $j$ should be a multiple of 32 so that 
\begin{equation}
N_B = 39j = 1248i, \quad i=1,2,3,\ldots
\end{equation}
Lastly, the size of the C transform is then determined:
\begin{equation}
N_C = \frac{32}{39} N_B.
\end{equation}
The dashed line in Figure \ref{fig:C2R_performance} shows the performance for candidate $N_B$s.  Unlike for 
the radix-2 cases, the configuration that has the best Gflop performance does not also provide the best Gsps
rate.  This is likely because the Gflop calculation is not an actual flop count, but a scaling from the 
original Cooley-Tukey algorithm \citep{cooley65}.  The Gsps metric is perhaps more useful in this case as it
shows that choices of $N_B <= 19968$ will have good performance.  However, the figure also demonstrates
that this non radix-2 FFT is about $2\times$ slower than the radix-2 case.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{aphids/C2R_performance.eps}
\caption{This figure shows the performance of radix-2, single precision, batched, complex-to-real, 1D transforms 
measured on our GPU cluster. The benchmark conventions follow FFTW  (http://www.fftw.org/speed/method.html).}
\label{fig:C2R_performance}
\end{figure}

>>>>

To enable VLBI observations with heteregeneous arrays, it is sometimes necessary to pre-process
the data before correlating.  In this chapter we describe a software solution to change the 
samping rate of the data taken at the SMA for the EHT's March 2015 observing campaign.  We 
focus on the sampling rate conversion problem and investigate several possible algorithms.  The project
has applications for future VLBI observations at the SMA as well as other sites with similar 
challenges.

@@

Very Long Baseline Interferometry (VLBI) uses similar techniques as the connected element arrays such as the 
SMA and ALMA
featured in the previous chapters.  By using detectors located thousands of miles away, perhaps on different 
continents or even in space, this technique can achieve milliarcsecond resolution at radio wavelengths.  Each 
station must have its own clock and, after the observation campaign, the recorded signals is correlated off-site. 
Examples of VLBI applications include geodesy, astrometry, and imaging \citep{felli89,sovers98}.  Connected
elements arrays may also participate in these experiments as a single, phased array site \citep{thompson01}.

Technical developments in the field continue to open new scientific opportunities.  One such exciting example 
is the development of mm and sub-mm wavelength VLBI by collaborations such as the Event Horizon
Telescope (EHT).  The EHT observes nearby super massive black holes on scales of their Schwarzschild radii and
their observations provide a unique lab to study general relativity and accretion physics.  One of the 
challenges that this experiment faces is the inhomogeneity of its array which is composed of many independently 
operated observatories.  While some of these telescopes use the same back-end hardware such as the 
state-of-the-art ROACH2 VLBI Digital Back-end \citep[R2DBE;][]{vertatschitsch16}, a few sites 
cannot use these systems and produce very 
different data products. In particular, the sampling rate of the phased ALMA, SMA, and CARMA arrays are 
incompatible with the standard VLBI rate.  Before the EHT's software correlator
\citep[DIFX;][]{deller07} can find fringes using these data, they must be resampled to the R2DBE rate of 
4096\,MHz.

This chapter summarizes the Adaptive Phased-array and Heterogeneous 
Interpolator, and Downsampler for SWARM (APHIDS), a resampling software designed for 
phased SMA data products.  We concentrate our discussion to early development and design in the project with 
a particular emphasis on the resampling problem.  APHIDS also has very general applications for 
post-processing of VLBI data since it features fast I/O, a modular threaded pipeline 
and Graphics Processor Unit (GPU) accelerated data processing.  The code base is open-source and available 
online at \url{https://github.com/sma-wideband/sdbe}.

The purpose of APHIDS is to take data recorded at the SMA and process it for correlation using
DIFX.  The SMA is a key site for the EHT that provides important East-West baselines and a near zero baseline 
with its neighbor the JCMT (see Figure \ref{fig:eht_baselines}).  The phased sum is calculated by the SMA correlator, the SMA Wideband Astronomical 
ROACH2 Machine (SWARM\footnote{\url{https://www.cfa.harvard.edu/twiki5/view/SMAwideband}}),
which is currently being deployed in stages as the array's bandwidth capabilities are 
upgraded.  For EHT observations taken in March 2015, SWARM recorded at a sampling rate of 2496\,MHz in 
two separate phased sums.

In \S\ref{sec:src_basics} we discuss the sampling rate conversion problem and several common solutions.
The current APHIDS resampling algorithm is described in \S\ref{sec:aphids} along with several important
design considerations.  We include measurements of performance as well as a successful
demonstration of the software's operation.  We end by summarizing the goals for near and long-term project 
development.

@@

Sample rate conversion, or resampling, is the process of taking a digital signal ($x[i]$) sampled at some rate 
($f_0$) and calculating new samples ($y[i]$) at a different rate ($f_1$).  To analyze this problem we use the 
ratio between the two rates: $f_1/f_0 = L/M$, where $L$ and $M$ are relatively prime integers.  A sample rate 
conversion can be helpfully thought of as the combination of upsampling and downsampling 
\citep{oppenheim10,lyons11}.  
Upsampling, (also commonly called interpolation or expansion), increases the sampling rate 
by a factor of $L$ with the insertion of $L-1$ zeroes between the original $x[i]$.  A low-pass, 
anti-imaging filter smooths the signal and supresses the high frequency spectral images greater than the original 
$f_0$ that have been introduced by the zero inserts.  Figure \ref{fig:upsampling} sketches this process in the 
time and frequency domain for when $L$ equals 3.

@@

\begin{figure}[t]
\centering
\begin{tikzpicture}
   \node[dspnodeopen,dsp/label=left]  (c0) {$x[i]$};
   \node[dspsquare,right=of c0]                     (c1) {\upsamplertext{L}};
   \node[dspfilter,right=of c1]                     (c2) {$H(\z)$};
   \node[dspsquare,right=of c2]                    (c3) {\downsamplertext{M}};
   \node[dspnodeopen,right=of c3,dsp/label=right]  (c4) {$y[j]$};
   \foreach \i [evaluate = \i as \j using int(\i+1)] in {0,1,...,3}
       \draw[dspconn] (c\i) -- (c\j);
\end{tikzpicture}
\caption{Basic multirate resampling signal flow graph.}
\label{fig:resample_basic}
\end{figure}

There have been recent development of polyphase filter representations for GPUs
\citep[i.e.][]{vanderveldt12,adamek14,kim14a}.  These filter structures can operate at the low sample rate of 
$f_0/M$
by introducing delays to switch the expander and decimatator \citep{crochiere81}.  However, they require 
some filter design and careful book-keeping or buffering to keep track of sample indices \citep{wang01}.
The implementation in \citet{kim14a} assigned each thread to a filter coefficient, parallelizing the 
inner product operation.  However, \citet{kim14a} found that their GPU kernel was dominated by the 
indexing operations.  \citet{adamek14} (building on the work by \citet{vanderveldt12}) presented an
implementation of a polyphase filter bank that could run at data rates in excess of 6.5\,GB/s 
(their estimate for the single channel output of the SKA's Low Frequency Aperture Array).

%  https://github.com/wesarmour/astro-accelerate
\begin{figure}[t!]
\epsscale{1.0}
\plotone{aphids/windows.eps}
\caption{Frequency response for FIR filters equivalent to linear and nearest interpolation when $L/M = 64/39$.
The orange region shows the first Nyquist zone of the target $L/Mf_0$ sample rate.  Spectral components at all 
other frequencies are aliased into this region.}
\label{fig:windows}
\end{figure}

For testing and development we also considered two elementary options for the resampling: linear and nearest 
neighbor interpolation.  Linear interpolation is:
\begin{equation}
y[i] = x[j] + (x[j+1] - x[j]) (if_0/f_1 - j)
\end{equation}
where $j = \mathrm{floor}(if_0/f_1)$.  This is equivalent to applying a ``tent'' FIR filter with $2L$ taps on the 
upsampled $Lf_0$ time series before decimation \citep{oppenheim10}.  Similarly, a $2L$ tap box-car filter is 
equivalent to nearest-neighbor interpolation:
\begin{equation}
y[i] = x[\mathrm{round}(if_0/f_1)].
\end{equation}

@@

\subsection{Resampling in the Fourier Domain} \label{sec:fourier_resampling}
In contrast to the previous methods, one can also implement a rational $L/M$ sample rate conversion entirely in 
the Fourier domain \citep{gold69}.  After accumulating $kM$ samples at a clock rate of $f_0$ 
(where $k=1,2,3\ldots$), the DFT 
returns spectral components spaced at $f_0/kM$.  If $f_1 > f_0$, the resampled spectrum is generated by inserting
$p$ zeros to match the new $f_1$ while maintaining the correct frequency components: 
\begin{equation}
\frac{f_1}{kM+p} = \frac{f_0}{kM}.
\end{equation} 
Solving for $p = k(L - M)$.  If $f_0 < f_1$, the spectrum is instead trimmed by $p$ samples.  The time series
sampled at $f_1$ can then be constructed from the inverse DFT.  This method is equivalent to sinc 
interpolation using an ideal low-pass filter and is a perfect interpolator
if the original, continuous signal has a discrete spectral density distribution that is band-limited below the 
Nyquist limit.  In practice, errors may be introduced from aliased spectral leakage.

Trimming or padding the spectrum is equivalent to convolving a normalized sinc function with the original series 
wrapped periodically.  Consequenty, errors will be introduced into the beginning and end of the
resampled signal.  How far and significantly this error propagates away from the time series' edges depends 
on the width of the sinc function which is set by the resampling factors $L$ and $M$: 
\begin{equation} \label{eq:sinc}
B\,\mathrm{sinc}(B t) \overset{\mathcal{F}}{\Longleftrightarrow} \begin{cases} 1 \quad |f| < B/2 \\ 0 \quad |f| > B/2 \end{cases}
\end{equation}
where $B = \mathrm{min}(f_0,L/Mf_0)$.  
The number of samples for which this makes a large difference is small and can be roughly treated as less than a 
1\% effect when $N > 100\,\mathrm{max}(L/M,1)$.  One option to mitigate this error to stitch together 
overlapping, resampled segments \citep{bi11} which requires more computation and memory.  A second strategy is to 
multiply the original signal by some tapered window function \citep{fraser89}.

For post-processing phased SWARM, DFT resampling is a good fit because one can access large chunks of 
the time series without accumulators in hardware and it has well-behaved errors
that can be easily modeled.  However, the speed of the FFT depends on the resampling factors ($L/M = 32/39$ for 
the March 2015 EHT observations) and zero-padding in the case that the $L\gg M$ may require large amounts of 
memory.  In practice, we also found that the memory space of the GPU could be a limiting factor not only 
on how much data could be simultaneously resampled, but also storing FFT plans (which depends on
values of $L$ and $M$).

@@

\begin{figure}[t!]
\linespread{1.}\selectfont{}
\begin{center}
\begin{tikzpicture} [node distance = 5cm, auto]
\node [cpu] (mark6_in) {Mark6 Reader};
\node[gpu,below of=mark6_in](reader){Reader};
\node[gpu,right of=reader](resampler){Resampler};
\node[cpu,right of=resampler](VDIF_pkt){VDIF packet \\ \& header};
\node[cpu,below of=VDIF_pkt](mark6_out){Mark6 Writer};

\path[line] (mark6_in) -- node[align=center] {BENG @ 2496 GHz \\ quantized to 2-bits}(reader);
\path[line] (reader) -- node[align=center] {spectra of \\ 16k complex64} (resampler);
\path[line] (resampler) --  node[align=center] {time series \\ @ 2048 GHz}(VDIF_pkt);
\path[line] (VDIF_pkt) -- node[align=center, left]{VDIF @ 4096 GHz \\ quantized to 2-bits}(mark6_out);
\end{tikzpicture}
\caption{This signal flow chart shows an overview of APHIDS.  The red boxes represent operations on the GPU 
while blue boxes are for the CPU.  Once the B-engine frames have been unpacked, each of the two phased sum 
streams are processed independently and in a serial fashion.}
\label{fig:aphids_flow_chart}
\end{center}
\end{figure}

@@

We are developing a software solution to the resampling problem called APHIDS.  This 
open-source project is written in C/CUDA using
the High Availibility Shared Pipeline 
(HASHPIPE)\footnote{\url{https://github.com/david-macmahon/hashpipe}}.  The software is designed to use 
modularized, lightweight threads to handle reading data packets from hard drives, resampling on a GPU cluster, 
and then writing back to separate disks.  These three steps are each loosely mapped to threads we call the 
vdif\_in\_net thread, vdif\_inout\_gpu\_thread, and vdif\_out\_net thread. 
The ultimate goal of is to process data streamed directly at the telescope.  The flowchart in Figure 
\ref{fig:aphids_flow_chart} shows how data progresses 
through the pipeline and this section will describe each block covering the major hardware components, software 
design, and data formats.  We will conclude with a discussion of future development and applications of APHIDS.

@@

APHIDS features three main hardware components: an I/O data system, CPU, and GPU cluster.  For I/O, the 
EHT observations taken by SWARM in March 2015 have been stored on four disk-packs that each hold eight, 
6\,TB hard drives (totaling 192\,TB of disk space).  The Mark6 VLBI data system has four modules 
that can hold one disk-pack and we use
two seprate Mark6 units to handle the recording and playback of data.  APHIDS itself is run on a separate CPU 
that controls the GPU cluster.

The GPU cluster has four GTX 980 GPUs, each with 4GB of local memory.  Data from the CPU (or host) is copied over 
a PCI-E bus into the local GPU (or device) memory.  Once the data is transferred, the host directs the device
to perform functions (or kernels) on the data which must be again moved from the local device memory onto 
on-chip memory for the individual streaming multiprocessors that make up the GPU.  After computation is done, 
the host transfers the data off of the device memory.

The observations were recorded using a scatter/gather file system which disperses the 
individual scan data over the four modules in a Mark6.  APHIDS uses fast, general purpose software called 
sgcomm (Scatter/Gather Communication) to read in 
the scattered data into CPU shared memory over a 10 Gigabit Ethernet.  Data from phased SWARM was 
streamed through a specialized version of the R2DBE called the SDBE (Swarm Digital Back End) and packaged into
a customized B-engine format.  Each B-engine frame is composed of 2048, 1056B VDIF frames 
\footnote{\url{http://www.vlbi.org/vdif/}} with 128 snapshots of a 
16384 sample SWARM spectra.  However, the current SDBE bitcode packs
neither the snapshots contiguously in time nor the spectrum samples contiguously in frequency.  Furthermore, 
each B-frames overlaps the preceding one in memory.  Our resampling algorithm handles $19968$\,SWARM samples
simultaneously, so in order to reduce the amount of book-keeping the inout thread buffers 40 B-engine frames to 
produce $39 \times 128$ complete SWARM snapshots.  The inout thread does preliminary checks to ensure that the 
B-engine frames are complete before assigning each buffer to a GPU.  The data is written as a time series using 
following the VDIF specification.

As mentioned previously, the data is split into two phased sums that together cover the desired 4096\,MHz of 
observing bandwidth.  Since DIFX can handle the chunks independently we do not stitch the spectra together. 
However, the observation was designed to include guard bands which must be removed by APHIDS 
(see Figure \ref{fig:swarm_amp_spec} and \S\ref{sec:resamp_block}).

\begin{figure}[t!]
\linespread{1.}\selectfont{}
\centering
\begin{tikzpicture}[node distance = 2.2cm, auto]
\node[align=center](init){};
\node[resamp,below of=init](irfft){(A) batched real IFFT \\ dimension of $32768$};
\node[resamp,below of=irfft](rfft){(B) batched real FFT \\ dimension of $39 \times 512$};
\node[resamp,below of=rfft](trim){Trim band and shift frequency};
\node[resamp,below of=trim](rifft){(C) batched real IFFT \\ dimension of $32 \times 512$};
\node[align=center,below of=rifft](final){};

\path[line](init) -- node[align=left]{ complex spectra \\ @ 2496\,MHz} (irfft);
\path[line](irfft) -- (rfft);
\path[line](rfft) -- (trim);
\path[line](trim) -- (rifft);
\path[line](rifft) -- node[align=left]{ time series \\ @ 2048\,MHz}(final);
\end{tikzpicture}
\caption{The resampling block trims the phased SWARM guard bands and shifts the passband to DC as part of the 
resampling operation.}
\label{fig:resampling_block}
\end{figure}

@@

The APHIDS resampling block is composed of three 1-dimensional FFTs (A, B, and C; see Figure 
\ref{fig:resampling_block}).  Following the cuFFT API reference to improve performace, we utilize batched
FFTs (launching more than one FFT at a time using the cuFFT API), out-of-place transforms, 
and operate on contiguous chunks of memory.  We also use single precision transforms to 
reduce the memory transfer bandwidth and ensure that 
the resampling block can be run on a single device.  The latter point is important since it allows the inout
thread to parallelize the resampling operation across the cluster and provide a roughly x4 speedup.

The first FFT (A) takes the complex valued, 16384 sample SWARM spectrum into the time domain, returning 
32768 real samples.  Since SWARM omits the Nyquist sample which is outside the SWARM guard band cutoff frequency, 
we insert a zero as the 16385th sample.  Since this transform is radix-2, the performance should benefit from 
the Cooley-Tukey algorithm.  However, on our cluster the cuFFT library exhibits linear improvement in Gflops 
with transform size up to N of 16384 (see Figure \ref{fig:C2R_performance}).  This behavior is consistent with 
the CUDA 7.0 performance 
report\footnote{\url{http://on-demand.gputechconf.com/gtc/2015/webinar/gtc-express-cuda7-performance-overview.pdf}} 
and we find that this transform, just ouside of the optimal regime, operates at 6.7 Gsps.

\begin{figure}[t!]
\epsscale{0.95}
\plotone{aphids/C2R_performance.eps}
\caption{This figure shows the performance of radix-2, single precision, batched, complex-to-real, 1D transforms 
measured on our GPU cluster (solid lines). The dashed lines show the same measures for non radix-2, candidate 
resampling FFT sizes ($N_B$).  The benchmark conventions follow FFTW  (http://www.fftw.org/speed/method.html).}
\label{fig:C2R_performance}
\end{figure}

@@

The choice of FFT size was set to maximize performance.  In order for the target sampling rate to be achieved
$N_B$ must be a multiple of 39.  Furthermore, so that the 150\,MHz guard band can be removed, we must ensure that 
150 is a multiple of $2496/N_B$:
\begin{equation}
\frac{150\times39}{2496} = \frac{75}{32} = \frac{i}{j}, \quad i,j = 1, 2,3, \ldots
\end{equation}
To ensure this, $j$ should be a multiple of 32 so that 
\begin{equation}
N_B = 39j = 1248i, \quad i=1,2,3,\ldots
\end{equation}
Lastly, the size of the C transform is then determined:
\begin{equation}
N_C = \frac{32}{39} N_B.
\end{equation}
The dashed line in Figure \ref{fig:C2R_performance} shows the performance for candidate $N_B$s.  Unlike for 
the radix-2 cases, the configuration that has the best Gflop performance does not also provide the best Gsps
rate.  This is likely because the Gflop calculation is not an actual flop count, but a scaling from the 
original Cooley-Tukey algorithm \citep{cooley65}.  The Gsps metric is perhaps more useful in this case as it
shows that choices of $N_B \leq 19968$ will have good performance.  However, the figure also demonstrates
that this non radix-2 FFT is about $2\times$ slower than the radix-2 case.
